---
title: "Heuristica"
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file and then run: -->
<!-- library(knitr); knit("README.Rmd") -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

This R package implements [heuristic](http://en.wikipedia.org/wiki/Heuristic) decision models, such as a [unit-weighted linear model](http://en.wikipedia.org/wiki/Unit-weighted_regression) and Gigerenzer and Goldstein's [Take The Best](http://en.wikipedia.org/wiki/Take-the-best_heuristic) (TTB), which uses just one cue to make its inference.  The models are designed for two-alternative choice tasks, such as which of two cities has a greater population given that one has a university and the other does not.  It also wraps more well-known models like regression and logistic regression into the two-alternative choice framework for easy comparison.  It provides functions to measure accuracy and robustness (out-of-sample accuracy).

The goal is to make it easy to explore the wide range of conditions in which simple heuristics can be more robust than more complex models.  Optimizing is not always better!

# The Task

This package is focused on two-alternative choice tasks, e.g. given two schools, which has a higher drop-out rate.  The output is categorical, not quantitative.

# A Simple Example

Here is a subset of data on Chicago public high school drop-out rates.
```{r}
schools <- data.frame(Name=c("Bowen", "Collins", "Fenger", "Juarez", "Young"), Dropout_Rate=c(25.5, 11.8, 28.7, 21.6, 4.5), Low_Income_Students=c(82.5, 88.8, 63.2, 84.5, 30.3), Limited_English_Students=c(11.4, 0.1, 0, 28.3, 0.1))
schools
```

## Fitting

To fit Take The Best-- ttbModel-- and a wrapped version of R's "lm" function for a linear regression-- regModel-- we specify the data set and the columns to use.  The 2nd column, `Dropout_Rate`, is the __criterion__ to be predicted.  The __cues__ are the following columns, indexes 3 and 4, percent of `Low_Income_Students` and percent of `Limited_English_Students`.

```{r}
ttb <- ttbModel(schools, 2, c(3:4))
reg <- regModel(schools, 2, c(3:4))
```

What does the fit look like?  We can examine Take The Best's cue validities and the regression coefficients.
```{r}
ttb$cue_validities
coef(reg)
```
Both Take The Best and regression give a higher weight to `Low_Income_Students` than `Limited_English_Students`, although of course how they use the weights differs.  Take The Best will use a lexicographic order, making its prediction based solely on `Low_Income_Students` as long as the schools have differing values-- which they do for all 5 schools in this data set.  That means it will ignore `Limited_English_Students` when predicting on this data set.  In contrast, regression will use a weighted sum of both cues, but with the most important cues weighted more.  

## Predicting

To see a model's predictions, we give it two rows and the fitted model to the __predictPair__ function.  It outputs 1 when it selects the first row passed to it and -1 when it selects the second row passed to it.  (We use the package's __oneRow__ helper function to select a row.)  Below we see that between the first and 2nd row, ttb predicts the first row, outputting 1.  Between the first and third row, it outputs -1, predicting the 2nd row passed to it (row 3) is greater.
```{r}
predictPair(oneRow(schools, 1), oneRow(schools, 2), ttb)
predictPair(oneRow(schools, 1), oneRow(schools, 3), ttb)
```

Interestingly, regression predicts exactly the opposite for these row pairs!
```{r}
predictPair(oneRow(schools, 1), oneRow(schools, 2), reg)
predictPair(oneRow(schools, 1), oneRow(schools, 3), reg)
```

Looking at the data set, we see that Take The Best was correct about row 1 (Bowen) vs row 2 (Collins) because Bowen has the higher `Dropout_Rate`.  It was also correct about row 1 (Bowen) vs. row 3 (Fenger) because Fenger has the higher `Dropout_rate`.  Regression was wrong on both.
```{r}
schools[c(1:3), c(1,2)]
```

## Assessing Overall Performance

For an overall measure of performance, we can measure the percent of correct inferences for all pairs of schools in the data with __percentCorrect__.  The function also lets us readily compare ttb and regression.
```{r}
percentCorrect(list(ttb, reg), schools)
```

Take The Best got 60% correct and regression got 50% correct, which is the same as chance.

Regression is the best linear unbiased model for the data.  But this data had a very small sample size of just 5 schools, and good estimates require more data.

This is an unusual case where TTB actually beat regression in a fitting task.  Usually ttb only wins in out-of-sample performance, e.g. fitting 5 schools and then predicting on other schools not used in the fit.

For a more realistic example, see the vignette with cross-validated out-of-sample performance on a complete data set.

# Package Installation

Uncomment and execute the line below to get the CRAN version:

```{r}
# install.packages("heuristica") 
```

Uncomment and execute the line below to get the development version.

```{r}
# Uncomment and execute the line below if you do not have devtools.
# install.packages("devtools") 
# devtools::install_github("jeanimal/heuristica")
# library("heuristica")
```

# Models

The package comes with the following models that implement predictPair.
* __ttbModel__: An implementation of [Take The Best](http://en.wikipedia.org/wiki/Take-the-best_heuristic). It sorts cues in order of cue validity, making a decision based on the first cue that discriminates (has differing values on the two objects).
* __ttbGreedyModel__: Take the Best using conditional cue validity (rather than cue validity).
* __unitWeightModel__: A [unit-weighted linear model](http://en.wikipedia.org/wiki/Unit-weighted_regression) that uses weights of +1 or -1 only.  An Exception is that a cue with no variance-- every value is the same-- gets a weight of 0.  Inspired by psychologist Robyn Dawes-- see citation below.
* __validityWeightModel__: A cue-validity-weighted linear model.  (In some publications, this was called franklinModel after Ben Franklin.)
* __regModel__: A regression model, a wrapper around R's lm to make it easier to compare with heuristics.  It generates a regression formula for you based on the matrix and column indices you give it.  It generates a prediction for each item in the pair-- e.g. estimates the population of Rostock and the population of Munich-- and then picks the item (city) with the higher estimate.
* __logRegModel__: A logistic regression model, a wrapper around R's glm.  This estimates the probability that Rostock is greater than Munich and chooses Rostock if the probability is greater than 50/50.

You can add your own models by also implementing a function related to __predictPair__, as described in a vignette.

# Data

The package comes with two data sets used in the book, Simple Heuristics That Make Us Smart.
* city_population (all binary cues)
* highschool_dropout

# Citations

Take The Best was first described in:
Gigerenzer, G. & Goldstein, D. G. (1996). "Reasoning the fast and frugal way: Models of bounded rationality". Psychological Review, 103, 650-669.

All of these heuristics were run on many data sets and analyzed in:
Gigerenzer, G., Todd, P. M., & the ABC Group (1999). [Simple heuristics that make us smart.](http://www.amazon.com/Simple-Heuristics-That-Make-Smart/dp/0195143817) New York: Oxford University Press. 

The research was also inspried by:
Dawes, Robyn M. (1979). "The robust beauty of improper linear models in decision making". American Psychologist, volume 34, pages 571-582. doi:10.1037/0003-066X.34.7.571 [archived pdf](http://www.cmu.edu/dietrich/sds/docs/dawes/the-robust-beauty-of-improper-linear-models-in-decision-making.pdf)


